The Galactic Midplane is Not a Plane: Implications for dynamical analysis with {\em Gaia} data and beyond

### 2019-04-22 paper/ms.tex
Change colors in cartoon image

Now suppose this observer makes a measurement of the true orbit at the blue
point or the pink point (i.e. at two different orbital phases). Then the blue
and pink lines correspond to the orbits that the observer would compute for
each point based on the potential model with the erroneous midplane. In action
space, this would correspond to a different value of $J_z$ for the blue and
pink points. In this way, assuming the wrong coordinate system induces a
phase-dependence in the actions estimated for the star, which in the correct
potential (in this example, the one with the correct midplane) should be
phase-independent.

\begin{figure*}
\plotone{fig/cartoon.pdf}
\caption{Cartoon approximation showing the effect an error in the determination of the
coordinate midplane can have on orbit integration and action estimation. The
$x$-axis shows the orbital phase and the $y$-axis the vertical height. The top
gray curve depicts an example ``true'' orbit oscillating about the true
midplane (horizontal solid gray line). Consider an observer who erroneously
assumes the midplane is located at the horizontal dashed line. Suppose this
observer measures the phase-space position of the star at two different
orbital phases (blue and pink points). If the observer were then to integrate
the star's orbit using a model potential with the erroneous midplane, they
would obtain the blue and pink curves for the star's orbit, respectively. The
actions estimated from these two  erroneous orbits would subsequently differ,
both from each other and from the actions estimated for the true orbit (in the
potential with the correct midplane). Hence an incorrect midplane in the
potential model assumed will induce phase-dependence in the actions estimated
for a given star in that potential.}
\label{fig:cartoon}
\end{figure*}

>>>>

Now suppose this observer makes a measurement of the true orbit at the purple
point or the green point (i.e. at two different orbital phases). Then the purple
and green lines correspond to the orbits that the observer would compute for
each point based on the potential model with the erroneous midplane. In action
space, this would correspond to a different value of $J_z$ for the blue and
pink points. In this way, assuming the wrong coordinate system induces a
phase-dependence in the actions estimated for the star, which in the correct
potential (in this example, the one with the correct midplane) should be
phase-independent.

\begin{figure*}
\plotone{fig/cartoon.pdf}
\caption{Cartoon approximation showing the effect an error in the determination of the
coordinate midplane can have on orbit integration and action estimation. The
$x$-axis shows the orbital phase and the $y$-axis the vertical height. The top
gray curve depicts an example ``true'' orbit oscillating about the true
midplane (horizontal solid gray line). Consider an observer who erroneously
assumes the midplane is located at the horizontal dashed line. Suppose this
observer measures the phase-space position of the star at two different
orbital phases (purple and green points). If the observer were then to integrate
the star's orbit using a model potential with the erroneous midplane, they
would obtain the purple and green curves for the star's orbit, respectively. The
actions estimated from these two  erroneous orbits would subsequently differ,
both from each other and from the actions estimated for the true orbit (in the
potential with the correct midplane). Hence an incorrect midplane in the
potential model assumed will induce phase-dependence in the actions estimated
for a given star in that potential.}
\label{fig:cartoon}
\end{figure*}
